---
title: "Medicaid Eligibility Across Chicago Neighborhoods"
format: html
editor: visual
echo: false
---

```{r}
packages_list <- 
  c("tigris", "tidycensus", "DT", "dplyr", "tidyr", "stringr", "sf", "devtools",
    "scales", "ggplot2", "plotly")
for (p in packages_list) {
  if (!p %in% installed.packages()[, "Package"]) install.packages(p)
  library(p, 
          character.only = TRUE, 
          quietly = TRUE)
}
```

```{r}
devtools::install_github("CMAP-REPOS/cmapgeo")

library(cmapgeo, 
        quietly = TRUE)
```



```{r}
census_key <- readLines(con = "census-key.txt")
Sys.setenv(census_api_key = census_key)
```

# Medicaid Rules

Illinois Legal Aid confirms that [Medicaid eligibility for adults in Illinois](https://www.illinoislegalaid.org/legal-information/am-i-eligible-medicaid) is 138% of the federal poverty level.

# Data Sources

The American Community Survey (ACS) 5-Year release has data down to the Census tract level, which each represent approximately 4,000 individuals. There are approximately 800 Census tracts in the city of Chicago.

[Table C17002](https://www.socialexplorer.com/data/ACS2021_5yr/metadata/?ds=ACS21_5yr&table=C17002) -- "Ratio Of Income To Poverty Level In The Past 12 Months" -- summarizes income to as fine as 25% intervals of income-to-poverty ratio. 

To estimate the number of adults that are Medicaid eligible by tract, we simply assume that `54%` of individuals with income reported between `125-149%` of the federal poverty line (FPL) to be below `138%` (FPL) and thus to be Medicaid eligible.[^`54%=(138-125)/(149-125)`]

```{r}
#| label: check metadata

meta <- 
  load_variables(year = 2021, dataset = "acs5") %>% 
  filter(str_detect(name, "C17")) %>% 
  data.frame() %>% 
  select(name, label) %>% 
  mutate(group = 
           case_when(str_detect(name, "1$")     ~ "Total",
                     str_detect(name, "[2-4]$") ~ "Elig",
                     str_detect(name, "5$")     ~ "Part",
                     TRUE                       ~ "Non-Elig"))

if (FALSE) {
  View(meta)
}
```

```{r}
#| label: pull census data for Cook county
pov <- 
  get_acs(geography = "tract",
          state = "IL", 
          county = "Cook",
          table = "C17002",
          year = 2021) %>% 
  select(variable, GEOID, estimate) %>% 
  merge(meta,
        by.x = c("variable"),
        by.y = c("name"),
        all.x = TRUE) %>% 
  mutate(elig_num = case_when(group == "Part" ~ estimate*0.5416,
                              group == "Non-Elig" ~ 0,
                              TRUE ~ estimate)) %>% 
  select(-variable)

if (FALSE) {
  pov %>% filter(GEOID == 17031010100) %>% select(label, group, estimate, elig_num)
}

```


```{r}
#| label: summarize and reshape

stats <- 
  pov %>% 
  filter(str_detect(group, "Elig|Part")) %>% 
  group_by(GEOID) %>% 
  summarize(elig = round(sum(elig_num))) %>% 
  merge(pov %>% filter(group == "Total") %>% select(GEOID, tot = elig_num),
        by = "GEOID") %>% 
  mutate(pct = elig / tot)
  
```

```{r}
#| label: merge poverty data into geographies and CCAs
stats_cca <- 
  merge(stats,
        xwalk_tract2cca %>% filter(pop_pct > 0.95),
        by.x = "GEOID",
        by.y = "geoid_tract",
        all.y = TRUE) %>% 
  merge(cca_sf %>% select(cca_num, cca_name, sqmi) %>% as.data.frame(),
        by = "cca_num") %>% 
  group_by(cca_name, sqmi) %>% 
  summarize(elig = sum(elig),
            tot = sum(tot)) %>% 
  mutate(pct = elig / tot,
         elig_sqmi = elig / sqmi / 1000)
```

# Medicaid Across Chicago Neighborhoods

```{r}
#| label: tbl-ranks
datatable(stats_cca %>% select(cca_name, elig_sqmi, pct), 
          colnames = c("Community Area" = "cca_name",
                       "# Eligible (thousands per sq mi)" = "elig_sqmi",
                       "% Eligible" = "pct"),
          rownames = FALSE) %>% 
  formatPercentage(columns = "% Eligible") %>% 
  formatRound(columns = c("# Eligible (thousands per sq mi)"))
  
```

```{r}
#| label: fig-scatter

my_scat <- 
  stats_cca %>% 
  ggplot(aes(x = pct,
             y = elig_sqmi,
             color = cca_name)) +
  geom_smooth(method = "lm",
              se = FALSE, color = "black") + 
  geom_point() +
  geom_text(data = stats_cca %>% filter(str_detect(cca_name, "Hyde")),
             aes(x = pct,
                 y = elig_sqmi,
                 label = cca_name),
            hjust = 0.5,
            vjust = +2) +
  scale_color_manual(breaks = c("Hyde Park"),
                     values = c("red", rep("gray", 76))) +
  scale_x_continuous(labels = percent, 
                     limits = c(0, NA)) +
  labs(title = "Estimated Medicaid Eligibility by Chicago Community Area",
       subtitle = "Counts (by area) vs Percent Eligibility",
       x = "% Eligible",
       y = "# Eligible (thousands per sq mi)",
       caption = "Source: American Community Survey 5-Year Data, 2021, Table C17002, with author's calculations.") +
  theme_minimal() +
  theme(legend.position = "none")
  
ggplotly(my_scat,
         tooltip = "color")

```
